\documentclass[11pt]{amsart}
\usepackage[margin=1in]{geometry}
\usepackage{paralist}
\usepackage{graphicx}
\usepackage{mathtools}

\theoremstyle{definition}
\newtheorem{exercise}{Exercise}
\newtheorem*{theorem}{Theorem}
\newtheorem*{definition}{Definition}

\begin{document}
\title{Linear Algebra}
\author{Strang, Section 4.2}
\maketitle

\section{The assignment}
\begin{compactitem}
\item Read section 4.2 of Strang (pages 206-214).
\item Read the following and complete the exercises below.
\end{compactitem}


\section{Orthogonal Projection}

One good use of the geometry in $\mathbb{R}^n$ is the concept of orthogonal projection. The basic idea is to mimic the behavior of shadows under sunlight. Our everyday experience leads us to thinking about the projection of a vector onto a plane (the ground--its roughly a plane), but if you imagine holding out a pencil you can summon up the visual of projection onto a line, too.

The key concept is to use the basic condition of orthogonality ($u \cdot v = 0$) to figure things out.

Note that everything in this section is done by projecting onto \emph{subspaces}! This is a bit of a restriction. In practice, this restriction can be removed by translating your whole problem to have a new origin.

\section{Sage instructions}

I have made a Sage worksheet file with some basic commands that you might find useful in investigating matrices. The file is called \texttt{section4\_2.sagews}.


\section{Questions for Section 4.2}
\setcounter{exercise}{122}

\begin{exercise} Find the projection matrix which computes projections of vectors in $\mathbb{R}^2$ onto the line $3x+2y=0$. (Since it goes through zero, it is a subspace.)

Find the orthogonal projection of the vector $\left( 17,3 \right)$ onto this line.
\end{exercise}

\begin{exercise}
Find the projection matrix which computes projections of vectors in $\mathbb{R}^3$ onto the line which is the intersection of the planes $x-2y+3z = 0$ and $y+2z=0$. (Again, that is a subspace.)

Find the orthogonal projection of the vector $\left(1,1,1\right)$ onto this line.
\end{exercise}

\begin{exercise}
Find the projection matrix which computes projections of vectors in $\mathbb{R}^3$ onto the plane $-2x + y +3z = 0$.

Find the orthogonal projection of the vector $\left( 9,7,-5\right)$ onto this plane.
\end{exercise}

\begin{exercise}
Find the projection matrix which computes projections of vectors in $\mathbb{R}^4$ onto the plane which is the intersectoin of $5x+y +w=0$ and $z+y+z+w=0$. (This subspace is the 2 dimensional plane where these two 3-dimensional hyperplanes meet.)

Find the orthogonal projection of the vector $\left(-3,1,-3,1\right)$ on this plane.
\end{exercise}


\end{document}




%sagemathcloud={"zoom_width":100}